Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2018-08-22 workflow/workflow.Rmd
added paragraph to Further Developments on downstream lineage DE methods.

---
title: "Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference"
author:

  - name: Fanny Perraudeau
    affiliation: 
      - Whole Biome, Inc., San Francisco, CA
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
  - name: Kelly Street
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
  - name: Sandrine Dudoit
    affiliation: Division of Epidemiology and Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA
abstract: Novel single-cell transcriptome sequencing assays allow researchers to measure gene expression levels at the resolution of single cells and offer the unprecendented opportunity to investigate at the molecular level fundamental biological questions such as stem cell differentiation or the discovery and characterization of rare cell types. However, such assays raise challenging statistical and computational questions and require the development of novel methodology and software. Using stem cell differentiation in the mouse olfactory epithelium as a case study, this integrated workflow provides a step-by-step tutorial to the methodology and associated software for the following four main tasks:(1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and pseudotimes; and (4) differential expression analysis along lineages. 
keywords: single-cell, RNA-seq, normalization, dimensionality reduction, clustering, lineage inference, differential expression, workflow
bibliography: ref.bib
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
editor_options: 
  chunk_output_type: console
---

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
if(packageVersion("zinbwave")<"1.1.0") stop("must have current master version to avoid bugs")
if(packageVersion("clusterExperiment")<"1.5.0-9009") stop("must have current devel version to avoid bugs")
if(packageVersion("scone")<"1.2.0") stop("must have current master version to avoid bugs")

library(BiocStyle)
library(knitr)
knitr::opts_chunk$set(cache=FALSE, error=FALSE, message=FALSE, warning=FALSE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5)
```

@@

Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells. To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include normalization, dimensionality reduction, clustering, the inference of cell lineages and pseudotimes, and the identification of differentially expressed (DE) genes. While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline. This is all the more true with novel sequencing technologies that allow an increasing number of cells to be sequenced in each run. For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.

scRNA-seq low-level analysis workflows have already been developed, with useful methods for quality control (QC), exploratory data analysis (EDA), pre-processing, normalization, and visualization. The workflow described in [@Lun2016] and the package `scater` [@McCarthy2017] are such examples based on open-source R software packages from the Bioconductor Project [@Huber2015]. In these workflows, single-cell expression data are organized in objects of the `SingleCellExperiment` class allowing integrated analysis [@SingleCellExperiment]. However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps such as cell clustering and lineage inference.

Here, we propose an integrated workflow for downstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` Bioconductor package; and (4) DE analysis along lineages.

One of the cornerstones of Bioconductor is the use of common data containers, implemented as classes in the S4 object-oriented system of R [@huber2015orchestrating].
Throughout the workflow, we use a single object of the `SingleCellExperiment` class to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment. 
The `SingleCellExperiment` class [@SingleCellExperiment] is an extension of the `SummarizedExperiment` class [@SummarizedExperiment] specifically designed for single-cell data (see Figure 2 of [@huber2015orchestrating] for an overview of `SummarizedExperiment`).
The use of the `SingleCellExperiment` class allows us to seamlessly transition between the steps of this workflow, as well as to integrate this workflow with the upstream QC and EDA steps provided in [@Lun2016].

@@

The scRNA-seq dataset we use as a case study was generated to study the differentiation of HBC stem cells into different cell types present in the olfactory epithelium. To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. The expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it. Cells were then assigned to different lineages using a statistical analysis pipeline analogous to that in the present workflow. Finally, results were validated experimentally using in vivo lineage tracing. Details on data generation and statistical methods are available in [@Fletcher2017; @Risso2017; @Risso2018; @Street2017].

It was found that the first bifurcation produces two lineage trajectories: the first one (purple curve in Figure \@ref(fig:stemcelldiff)) generates mature sustentacular cells (mSUS) through a differentiation process characterized by the absence of cell cycle and division. The other lineage trajectory (blue and orange curves in Figure \@ref(fig:stemcelldiff)) generates proliferative GBCs, which are characterized by cell cycle and division. This, in turn, diverges into two lineages that generate mature olfactory sensory neurons (mOSN; orange curve) and microvillous cells (MV; blue curve), respectively.

In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes-by-cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.

Note that the pre-processing steps needed to produce such a count matrix from the raw FASTQ files are outside of the scope of this paper. These are very important steps that should be taken with care and the optimal workflow depends on the technology and the protocol used to collect the data (e.g., 3'-end sequencing, droplet-based methods versus full-length mRNA, plate-based methods). 
One way to carry out these steps within the Bioconductor framework is the `scPipe` package [@tian2017scpipe], which provides functions to quantify gene expression starting from raw sequencing files. More details can be found in the `scPipe` vignette and in [@tian2017scpipe].

## Package installation and versions

The source code to reproduce the analysis in the workflow is available at https://github.com/fperraudeau/singlecellworkflow. The packages used in the workflow can be installed using the following in R:

```{r, eval=FALSE}
# CRAN
install.packages(c("doParallel", "gam", "RColorBrewer"))

# Bioconductor
install.packages("BiocManager")
BiocManager::install(c("scone", "zinbwave", "clusterExperiment", "slingshot",
                       "BiocParallel", "DESeq2", "scater"))
```

Note that in order to successfully run the workflow, we need the following versions
of the Bioconductor packages: `scone (>=1.5.0)`, `zinbwave (>=1.3.2)`, and `clusterExperiment (>=2.1.3)`. We recommend running Bioconductor 3.8 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).

We are now ready to load the packages.

```{r packages}
library(BiocParallel)
library(clusterExperiment)
library(scone)
library(zinbwave)
library(slingshot)
library(doParallel)
library(gam)
library(RColorBrewer)

@@

## Parallel computing

To give users an idea of the time needed to run the workflow, the function `system.time` was used to report computation times for time-consuming functions. Computations were performed with 6 cores on a iMac with a 4 GHz Intel Core i7 processor and 32 GB of RAM. The `BiocParallel` package is used to allow for parallel computing in the `zinbwave` function. Users with a different operating system may change the package used for parallel computing and the `NCORES` variable below.

@@

urls = c(paste0("https://www.ncbi.nlm.nih.gov/geo/download/",
                "?acc=GSE95601&format=file&file=GSE95601",
                "%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz"),
         paste0("https://raw.githubusercontent.com/rufletch/",
                "p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt"))

@@

The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to any cluster in the original publication.

```{r}
# Extract QC metrics
qc <- as.matrix(protocolData(Cufflinks_eSet)@data)[,c(1:5, 10:18)]

@@

# Symbol for cells not assigned to a lineage in original data
metadata$publishedClusters[is.na(metadata$publishedClusters)] <- -2
```

As discussed above, we store the data in an object of class `SingleCellExperiment`. In addition to keeping the expression values and their associated metadata within a single object, `SingleCellExperiment` allows the user to store normalization factors and the results of dimensionality reduction methods (such as `zinbwave`) within the same object. Throughout the workflow, we will highlight when this special features of `SingleCellExperiment` are used.

```{r}
dataObj <- SingleCellExperiment(assays = list(counts = E),
                           colData = metadata)
dataObj
```

There are three special sets of genes in this dataset: the CreER gene corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details); [*** SD: Do you mean the "Sox2-eGFP gene"?] the Sox2-eGFP transgenic was used for FACS sorting; the ERCC spike-in sequences are synthetic genes spiked-in at known concentrations and used for quality control purposes.

We can specify spike-in genes within the `SingleCellExperiment` class. In this way, it is easier to retrieve their expression measures. In addition, some specific functions may make direct use of this information (see, e.g., the `computeSpikeFactors` function in the `scran` package).

In this case, we specify two different sets of spike-ins, the ERCC spike-ins and the transgenic genes (eGFP and Cre-ER). This is a slight abuse of terminology, as the CreER and eGFP genes are not exactly spike-ins, but it is nonetheless useful to specify that these are not endogenous genes.

```{r}
# Mark ERCC and CreER genes as spike-ins
isSpike(dataObj, "ERCC") <- grepl("^ERCC-", rownames(dataObj))
isSpike(dataObj, "Transgene") <- grepl("^CreER$", rownames(dataObj)) | 
                                  grepl("^eGFP$", rownames(dataObj)) 
```

We can access a specific set of spike-ins (or all of them) by using the `isSpike` function.

```{r}
table(isSpike(dataObj))
table(isSpike(dataObj, "Transgene"))
```

The total number of reads mapped to spike-ins is often a good quality control measure. Hence, we include that into our `colData`.

```{r}
ercc <- counts(dataObj[isSpike(dataObj, "ERCC"),])
colData(dataObj)$ERCC_reads <- colSums(ercc)
```

Note that here we use the function `counts()` from the `SingleCellExperiment` package that retrieves the matrix stored in the assay slot named `counts`. This is equivalent to `assay(dataObj, "counts")` but is obviously more convenient when accessing counts multiple times. It is hence a good idea to name the assay slot that stores the original counts "counts", as done above.

## Exploratory data analysis

Cells were processed in 18 different runs of the Fluidigm C1 system: we refer to each run as a "batch" and to the technical differences between each run as "batch effects". Note that, in this particular case, the batch structure is essentially defined by the limitation of the C1 system to process up to 96 cells at a time and it is hence natural to identify each C1 run as one batch. Batch effects are ubiquitous in high-throughput genomics [@leek2010tackling] and single-cell RNA-seq is not an exception [@hicks2017missing]. In other contexts, depending on experimental design, batches may correspond to different library preparations, different labs in which the data are processed, or different times at which samples are collected.

```{r batch}
batch <- colData(dataObj)$Batch
col_batch = c(brewer.pal(9, "Set1"), brewer.pal(8, "Dark2"), 
              brewer.pal(8, "Accent")[1])
names(col_batch) = unique(batch)
table(batch)
```

Often, different batches differ in terms of data quality. In particular, we observed that the number of reads mapping to spike-ins can differ dramatically across batches.

```{r spikebatch, fig.cap="Distribution of ERCC spike-in read counts per batch."}
boxplot(colData(dataObj)$ERCC_reads~batch, col = col_batch,
        outline = 0, ylab="Reads mapped to ERCC", las = 3)
```

Figure \@ref(fig:spikebatch) shows how for some batches no reads were mapped to ERCC spike-ins, presumably because the spike-ins were not included. Since the utility of the spike-ins is limited in this example, we remove the spike-ins from our object.

```{r}
# added chunk of code
library(scater)
sce <- calculateQCMetrics(
    dataObj,
    feature_controls = list(
        ERCC = isSpike(dataObj, "ERCC")
    )
)
cData <- colData(sce)
cData <- as.data.frame(cData)
ggplot(cData, 
       aes(x = total_counts, 
           y = total_counts_ERCC,
           col = batch)) + 
  geom_point()
```

```{r}
dataObj <- dataObj[!isSpike(dataObj),]
dataObj <- clearSpikes(dataObj)
```

In the original work [@Fletcher2017], cells were clustered into 14 different clusters, with 233 cells not assigned to any cluster (i.e., cluster label of `-2`).

```{r original}
publishedClusters <- colData(dataObj)[, "publishedClusters"]
col_clus <- c("transparent", "#1B9E77", "antiquewhite2", "cyan", "#E7298A", 
              "#A6CEE3", "#666666", "#E6AB02", "#FFED6F", "darkorchid2", 
              "#B3DE69", "#FF7F00", "#A6761D", "#1F78B4")
names(col_clus) <- sort(unique(publishedClusters))
table(publishedClusters)
```

@@

## Sample filtering

Using the Bioconductor package `scone` [@Cole235382], we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes. For each criterion, a sample is filtered out if it is below a user-defined hard threshold (which we leave at default value) or an adaptive threshold, defined in terms of number of standard deviations from the mean or median absolute deviation from the median.

```{r scone, fig.cap="SCONE: Filtering of low-quality cells."}
# QC-metric-based sample-filtering
data("housekeeping")
hk = rownames(dataObj)[toupper(rownames(dataObj)) %in% housekeeping$V1]

mfilt <- metric_sample_filter(counts(dataObj), 
                              nreads = colData(dataObj)$NREADS,
                              ralign = colData(dataObj)$RALIGN,
                              pos_controls = rownames(dataObj) %in% hk,
                              zcut = 3, mixture = FALSE,
                              plot = TRUE)
```

The main parameter to set in this function is `zcut`, which represents the number of standard deviations from the mean and median absolute deviation from the median to use as threshold. We empirically found that using a threshold of three works well for typical datasets.

See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.

```{r sconeFilt}
# Simplify to a single logical
mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
dataObj <- dataObj[, mfilt]
dim(dataObj)
```

After sample filtering, we are left with `r ncol(dataObj)` good quality cells.

## Identification of most variable genes

Often in unsupervised analyses of expression data, it is useful to focus on the genes that vary the most in expression across cells and are thus likely to be the most informative. Filtering out low-variance genes allows us to reduce the complexity of the problem, while retaining the most information in the dataset.

Here, for computational efficiency, we retain only the 1,000 most variable genes, following a simple `log(x+1)` transformation. This seems to be a reasonable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]). In general, however, we recommend care in selecting a gene filtering scheme, as an appropriate choice is dataset-dependent and many different strategies have been proposed in the literature.

To compute the gene-wise variances and keep only the most variable genes, we use two functions from the `clusterExperiment` package. In particular, `makeFilterStats` will compute the variance of the gene expression data with a given transformation applied to the counts (in this case `log(x+1)`) and store the variance values in the `rowData` slot of the object [*** SD: Which object?]. `filterData` will then use the computed values to keep only the desired number of genes.

```{r}
dataObj <- makeFilterStats(dataObj, filterStats = "var", transFun = log1p)
dataObj <- filterData(dataObj, percentile=1000, filterStats="var")
rowData(dataObj)
```

One drawback of this strategy is that the logarithm is not a variance stabilizing transformation. Variance stabilizing transformations based on the negative binomial distribution are available, e.g., in the `DESeq2` package, and may be a better choice in real data analyses. Note that the `makeFilterStats` function of `clusterExperiment` can be used with any user-supplied function. Here, we show an example with the `vst` function from `DESeq2`.

```{r, eval=FALSE}
## Not run
library(DESeq2)
dataObj <- makeFilterStats(dataObj, filterStats = "var", transFun = vst,
                           filterNames = "var_vst")
dataObj <- filterData(dataObj, percentile=1000, filterStats="var_vst")
```

Alternatively, the `scran` package [@Lun2016] includes the `trendVar` and `decomposeVar` functions, which estimate the mean-variance relationship and the technical and biological components of the variance, to select those genes with a high biological variance. See [@Lun2016] for details.

```{r}
dataObj
```

@@

In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime inference, and the identification of DE genes. This allows the data to become more tractable, both from a statistical (cf. curse of dimensionality) and computational point of view. Additionally, technical noise can be reduced while preserving the often intrinsically low-dimensional signal of interest [@Peer2017; @Pierson2015; @Risso2017].

Here, we perform dimensionality reduction using the zero-inflated negative binomial-based wanted variation extraction (ZINB-WaVE) method implemented in the Bioconductor package `zinbwave`. The method fits a ZINB model that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a cell-level intercept, which serves as a global-scaling normalization factor. The user can also specify both gene-level and cell-level covariates. The inclusion of cell-level covariates enables normalization for batch effects, while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects). A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema). For greater detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].

```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2018)."}
knitr::include_graphics("zinb_schema.png")
```

The model can be thought of as a generalization of a factor analysis model, in which the biological signal of interest is assumed to be explained by a small number of "factors", represented in the model by the matrix `W` (Figure \@ref(fig:zinbschema)).

As with most dimensionality reduction methods, the user needs to specify the number of factors (`K`, the number of columns of `W`), which correspond to the dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects via the matrix `X`. Note that if the user includes more genes in the analysis, it may be preferable to reduce `K` to achieve a similar computational time.

```{r zinb}
print(system.time(dataObj <- zinbwave(dataObj, K = 50, X = "~ Batch",
                                      residuals = TRUE,
                                      normalizedValues = TRUE)))
```

@@

The function `zinbwave` returns a `SingleCellExperiment` object that includes the normalized expression measures (`notmalizedValues`), defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that the low-dimensional matrix `W` is not included in the computation of the residuals to avoid the removal of the biological signal of interest.

```{r norm}
norm <- assays(dataObj)$normalizedValues
norm[1:3,1:3]
```

@@

```{r boxplotNorm, fig.cap="ZINB-WaVE: Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
batch <- colData(dataObj)$Batch
norm_order <- norm[, order(as.numeric(batch))]
col_order <- col_batch[batch[order(as.numeric(batch))]]
boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
        border = col_order, xaxt = "n", ylab="Expression measure")
abline(h=0)
```

In addition to the normalized values, the `zinbwave` function added several new components to the object: the assay `residuals` contains the residuals of the model that can be used to check the goodness-of-fit of the model. The `weights` assay contains observational weights that can be used to downweight the effect of zero counts on differential expression (see [@van2018observation] and the `zinbwave` vignette for more details).

```{r}
assayNames(dataObj)
```

We set the normalized values as our first assay, so that they become the default in the plots below.

```{r setNormAssay}
assays(dataObj) <- assays(dataObj)[c("normalizedValues", "counts", "weights",
                                     "residuals")]
```

>>>>

---
title: "Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
  - name: Kelly Street
    affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA

abstract: Novel single-cell transcriptome sequencing assays allow researchers to measure gene expression levels at the resolution of single cells and offer the unprecendented opportunity to investigate at the molecular level fundamental biological questions such as stem cell differentiation or the discovery and characterization of rare cell types. However, such assays raise challenging statistical and computational questions and require the development of novel methodology and software. Using stem cell differentiation in the mouse olfactory epithelium as a case study, this integrated workflow provides a step-by-step tutorial to the methodology and associated software for the following four main tasks:(1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and pseudotimes; and (4) differential expression analysis along lineages.

keywords: single-cell, RNA-seq, normalization, dimensionality reduction, clustering, lineage inference, differential expression, workflow

bibliography: ref.bib

vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
---

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
if(packageVersion("zinbwave")<"0.99.6") stop("must have current devel version to avoid bugs")
if(packageVersion("clusterExperiment")<"1.3.2") stop("must have current devel version to avoid bugs")
if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")

library(BiocStyle)
library(knitr)
knitr::opts_chunk$set(cache=TRUE, error=FALSE, message=FALSE, warning=FALSE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5)
```

@@

Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells. To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include normalization, dimensionality reduction, clustering, the inference of cell lineages and pseudotimes, and the identification of differentially expressed (DE) genes. While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline. This is all the more true, with novel sequencing technologies that allow an increasing number of cells to be sequenced in each run. For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.

scRNA-seq low-level analysis workflows have already been developed, with useful methods for quality control (QC), exploratory data analysis (EDA), pre-processing, normalization, and visualization. The workflow described in [@Lun2016] and the package `scater` [@McCarthy2017] are such examples based on open-source R software packages from the Bioconductor Project [@Huber2015]. In these workflows, single-cell expression data are organized in objects of the `SCESet` class allowing integrated analysis. However, these workflows are mostly used to prepare the data for further downstream analysis and do not focus on steps such as cell clustering and lineage inference.

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages. Throughout the workflow, we use a single `SummarizedExperiment` object to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment.

@@

The scRNA-seq dataset we use as a case study was generated to study the differentitation of HBC stem cells into different cell types present in the olfactory epithelium. To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. The expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it. Cells were then assigned to different lineages using a statistical analysis pipeline analogous to that in the present workflow. Finally, results were validated experimentally using in vivo lineage tracing. Details on data generation and statistical methods are available in [@Fletcher2017; @Risso2017; @Street2017].

It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs. Then, the GBC lineage, in turn, branches off to give rise to mOSN, microvillous (MV) cells, and cells of the Bowman gland (Figure \@ref(fig:stemcelldiff)). In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.

## Package versions

The following packages are needed.

```{r packages}
# Bioconductor
library(BiocParallel)
library(clusterExperiment)
library(scone)
library(zinbwave)

# GitHub
library(slingshot)

# CRAN
library(doParallel)
library(gam)
library(RColorBrewer)

@@

Note that in order to successfully run the workflow, we need the devel versions
of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.6)`, and `clusterExperiment (>=1.3.2)`. We recommend running Bioconductor 3.6 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).

## Parallel computating

To give an idea to the users of the time needed to run the workflow, function `system.time` was used to report computation times for the time consuming functions. Computations were performed with 2 cores on a MacBook Pro (early 2015) with a 2.7 GHz Intel Core i5 processor and 8 GB of RAM. The `BiocParallel` package is used to allow for parallel computing in `zinbwave` function. Users with a different operating system may change the package used for parallel computing and the `NCORES` variable below.

@@

urls = c("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz",
         "https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt")

@@

We remove the ERCC spike-in sequences and the CreER gene, as the latter corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details).

```{r}
# Remove ERCC and CreER genes
cre <- E["CreER",]
ercc <- E[grep("^ERCC-", rownames(E)),]
E <- E[grep("^ERCC-", rownames(E), invert = TRUE), ]
E <- E[-which(rownames(E)=="CreER"), ]
dim(E)
```

Throughout the workflow, we use the class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object. The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to any cluster in the original publication.

```{r}
# Extract QC metrics
qc <- as.matrix(protocolData(Cufflinks_eSet)@data)[,c(1:5, 10:18)]
qc <- cbind(qc, CreER = cre, ERCC_reads = colSums(ercc))

@@

# Symbol for cells not assigned to a lineage in original data
metadata$publishedClusters[is.na(metadata$publishedClusters)] <- -2

se <- SummarizedExperiment(assays = list(counts = E),
                           colData = metadata)
se
```

Using the Bioconductor R package `scone`, we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes. See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.

```{r scone, fig.cap="SCONE: Filtering of low-quality cells."}
# QC-metric-based sample-filtering
data("housekeeping")
hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]

mfilt <- metric_sample_filter(assay(se), 
                              nreads = colData(se)$NREADS,
                              ralign = colData(se)$RALIGN,
                              pos_controls = rownames(se) %in% hk,
                              zcut = 3, mixture = FALSE,
                              plot = TRUE)
```

```{r sconeFilt}
# Simplify to a single logical
mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
se <- se[, mfilt]
dim(se)
```

After sample filtering, we are left with `r ncol(se)` good quality cells.

Finally, for computational efficiency, we retain only the 1,000 most variable genes. This seems to be a reasonnable choice for the illustrative purpose of this workflow, as we are able to recover the biological signal found in the published analysis ([@Fletcher2017]). In general, however, we recommend care in selecting a gene filtering scheme, as an appropriate choice is dataset-dependent.

```{r}
# Filtering to top 1,000 most variable genes
vars <- rowVars(log1p(assay(se)))
names(vars) <- rownames(se)
vars <- sort(vars, decreasing = TRUE)
core <- se[names(vars)[1:1000],]
```

```{r}
core
```

Metadata for the cells are stored in the slot `colData` from the `SummarizedExperiment` object. Cells were processed in 18 different batches.

```{r batch}
batch <- colData(core)$Batch
col_batch = c(brewer.pal(9, "Set1"), brewer.pal(8, "Dark2"), 
              brewer.pal(8, "Accent")[1])
names(col_batch) = unique(batch)
table(batch)
```

In the original work [@Fletcher2017], cells were clustered into 14 different clusters, with 151 cells not assigned to any cluster (i.e., cluster label of `-2`).

```{r original}
publishedClusters <- colData(core)[, "publishedClusters"]
col_clus <- c("transparent", "#1B9E77", "antiquewhite2", "cyan", "#E7298A", 
              "#A6CEE3", "#666666", "#E6AB02", "#FFED6F", "darkorchid2", 
              "#B3DE69", "#FF7F00", "#A6761D", "#1F78B4")
names(col_clus) <- sort(unique(publishedClusters))
table(publishedClusters)
```

@@

@@

In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and the identification of DE genes. This allows the data to become more tractable, both from a statistical (cf. curse of dimensionality) and computational point of view. Additionally, technical noise can be reduced while preserving the often intrinsically low-dimensional signal of interest [@Peer2017; @Pierson2015; @Risso2017].

Here, we perform dimensionality reduction using the zero-inflated negative binomial-based wanted variation extraction (ZINB-WaVE) method implemented in the Bioconductor R package `zinbwave`. The method fits a ZINB model that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a cell-level intercept, which serves as a global-scaling normalization factor. The user can also specify both gene-level and cell-level covariates. The inclusion of observed and unobserved cell-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects). A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema). For greater detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].

```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
knitr::include_graphics("zinb_schema.png")
```

As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects via the matrix `X`. Note that if the users include more genes in the analysis, it may be preferable to reduce `K` to achieve a similar computational time.

```{r zinb}
print(system.time(se <- zinbwave(core, K = 50, X = "~ Batch",  
                                 residuals = TRUE,
                                 normalizedValues = TRUE)))
```

@@

The function `zinbwave` returns a `SummarizedExperiment` object that includes normalized expression measures, defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that, in this case, the low-dimensional matrix `W` is not included in the computation of residuals to avoid the removal of the biological signal of interest.

```{r norm}
norm <- assays(se)$normalizedValues
norm[1:3,1:3]
```

@@

```{r boxplotNorm, fig.cap="ZINB-WaVE: Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
norm_order <- norm[, order(as.numeric(batch))]
col_order <- col_batch[batch[order(as.numeric(batch))]]
boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
        border = col_order, xaxt = "n", ylab="Expression measure")
abline(h=0)
```

The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by the original clusters (Figure \@ref(fig:pcanorm)). Overall, it seems that normalization was effective at removing batch effects without removing biological signal, in spite of the partial nesting of batches within clusters.

```{r pcanorm, fig.cap="ZINB-WaVE: PCA of normalized expression measures, where each point represents a cell. Cells are color-coded by batch (left panel) and by original published clustering (right panel)."}
pca <- prcomp(t(norm))
par(mfrow = c(1,2))
plot(pca$x, col = col_batch[batch], pch = 20, main = "")
plot(pca$x, col = col_clus[as.character(publishedClusters)], pch = 20, main = "")
```
